Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-04-21 thesis/draft.tex
modified:   draft.tex

\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{subcaption}

@@

\lhead{Ningrui Xie}
\chead{Thesis First Draft}
\rhead{ID: 400323416}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0pt}
\newcommand{\perday}{\ensuremath{\mathrm{/day}}}

@@

\newpage
\section{Abstract}
Simple dynamical models based on Ordinary Differential Equations (ODE) are widely utilized to forecast the spread of infectious diseases and to analyze other biological systems. A limitation of this framework is its implicit assumption that residence time in any given state, such as the infectious state, follows an exponentially distributed pattern when a constant rate is employed. This assumption is often not realistic. A common solution is the application of the Linear Chain Trick (LCT), wherein a single stage is subdivided into multiple substages, each follows an identical exponential distribution, thereby yielding a more realistic Erlang distribution for the duration. However, the LCT poses challenges by restricting parametric flexibility, particularly resulting in less flexible dwell time distributions, and it often necessitates comparisons between models with varying structures (i.e. different numbers of substages). Here, we propose a Geometric Chain Trick (GCT) employing a fixed number of substages with transition rates forming a geometric sequence. Through mathematical analysis, simulation, and model fitting, we demonstrate that our approach retains the principal benefits of the LCT while providing increased flexibility and facilitating easier data fitting. Unlike the recently introduced Generalized Linear Chain Trick (GLCT), which incorporates higher-dimensional auxiliary stages to model delays in transitions between main compartments, our method offers a more direct means of achieving flexibility. Although this study focuses on the application of the LCT and GCT in the epidemic SIR (susceptible-infectious-recovered) model, the potential applications of our approach span a broad array of dynamical models, including those for infectious diseases and other systems (e.g. population dynamics, biochemical reaction networks), offering advantages in computational efficiency, ease of parameter estimation, and the flexibility of time distributions.

\section{Introduction}
Ordinary differential equations (ODEs) are extensively utilized for mathematical modeling in the biological sciences, offering a robust framework to accurately describe the temporal dynamics of systems. Their capacity to model continuous state changes makes them essential in capturing complex biological phenomena such as population dynamics \cite{salisbury2011mathematical}\cite{sego2021generation}, epidemiological patterns of infectious diseases \cite{Anderson1991}\cite{Diekmann2000}\cite{Feng2016}, and cellular processes like proliferation in oncogenesis \cite{baker1998modelling}\cite{jarrett2018mathematical}. The inherent flexibility and simplicity of ODEs make them not only easy to solve and understand but also easy to manipulate and accommodate to various situations.

In the field of epidemiology, ordinary differential equations (ODEs) serve as a crucial tool for characterizing disease progression and making predictions. During the COVID-19 pandemic, they were extensively utilized to guide policy-making, especially in disease control and public health interventions \cite{thompson2020epidemiological}. A commonly used model is the SIR compartmental model \cite{Anderson1991}\cite{kermack1927contribution}, which divides the population into three states: Susceptible $(S)$, Infectious $(I)$, and Recovered $(R)$. In the model, a susceptible individual, upon contact with the infected population at a rate $\beta$ $(\frac{1}{time})$, transitions immediately to the infectious state. An infected individual then moves to the recoverd state at a recovery rate of $\gamma$ $(\frac{1}{time})$. However, one major limitation of this model it its implicit assumption that the duration of the infectious stage follows an exponential distribution. This can be observed in a simplified equation that models the dynamics of the infected population. It assumes the entire population was infected at the outset, and individuals can only transition out of the infectious stage at a rate of $\gamma$:
\begin{align}
    \frac{dI}{dt} = -\gamma I
\end{align}
Solving this equation explicitly yields
\begin{align}
    I(t) = I_0 e^{-\gamma t}
\end{align}
where $I_0$ represent the initial infected population. Formula (2) implies a stochastic state transition model where the time spend in the infectious $I$ is exponentially distributed with a mean stage duration of $\frac{1}{\gamma}$. This dwell time distribution suggests that an individual has the highest probability of recovering from the disease immediately after infection, which is often not realistic. In reality, a delay between the successful infection of an individual and the recovery from the disease is inevitable. The work conducted by Byrne AW et al. \cite{byrne2020inferred} presents the simulation of parameter distribution inferred for the duration of the infectious period, and the shape of this distribution closely resembles a gamma distribution. The study by Greenhalgh and Day (2017) \cite{greenhalgh2017time}, utilizing measles outbreak data, also demonstrates that the recovery rate from an infectious disease is not static but can vary based on a wide range of factors.

It is important to note that even when the mean is constant, the shape of the stage duration distributions (i.e., the timing of stage transitions) can significantly impact the dynamics and the outcomes of predictive models in applied settings \cite{krylova2013effects}\cite{keeling2002understanding}\cite{wearing2005appropriate}\cite{nguyen2008noise}. Consequently, it becomes crucial to consider the type of distribution characterizing an individual in a specific state when constructing these models. A standard approach for developing models that incorporate different stage duration distributions is through the use of integro-differential equations (IDEs) or integral equations (IEs) \cite{hurtado2019generalizations}\cite{kermack1927contribution}\cite{hethcote1980integral}. IDEs facilitate the extension of the SIR model to include arbitrary distributions for the duration of infectiousness, as discussed in various sources \cite{feng2000endemic}\cite{hethcote1980integral}\cite{ma2006generality}. This approach allows for the capture of the 'memory effect', where the current rate of change in each compartment of the model (Susceptible, Infectious, Recovered) is influenced by the entire history of the epidemic, rather than relying on a constant rate assumption.

@@

When applying LCT into the SIR model, the number of substages $n$ determines the shape of the stage duration distribution. As $n$ increases, the variance decreases, resulting in a distribution that is more peaked and narrow around the mean value $\frac{1}{\gamma}$. As $n$ approaches infinity, the distribution converges to a delta function (a deterministic value, implying that all individuals who become infectious at time $t$ recover at exactly time $t + \frac{1}{\gamma}$), and the system evolves into a delay differential equation \cite{krylova2013effects}\cite{hethcote1980integral}.

The Linear Chain Trick (LCT) is extensively employed in modeling biological processes, not only in quantifying disease progression \cite{lloyd2001destabilization}\cite{lloyd2001realistic}, but also in analyzing virus \cite{lloyd2001dependence}\cite{kakizoe2015method} and population \cite{cushing2013integrodifferential} dynamics. Studies have demonstrated that switching the stage duration distribution from an exponential to an Erlang distribution leads to predictions that align more closely with real data points \cite{kakizoe2015method}. However, despite the increased realism they bring to modeling disease processes, employing LCT in ODE models presents several limitations. One primary challenge is the inflexibility of the model's parameters. To integrate independent exponentially distributed random variables into the ODE system, $n$ must be an integer. As mentioned above, since the shape of the stage duration distribution is determined by $n$, it would also be constrained to discrete changes. Another inherent limitation of using LCT is the cumbersome and time-consuming process of determining the appropriate number of substage to divide. Multiple models with varying substages must be constructed and subjected to repeated data-fitting processes to identify the optimized parameter value. Prior research has investigated approaches for estimating the number of states in the linear chain, employing techniques such as profile likelihood \cite{raue2009structural} and model reduction \cite{maiwald2016driving}. Yet, a comparative study of these methods indicates that the typical quality of experimental data is insufficient for reliably estimating the chain length using these approaches \cite{hauber2020estimating}. These limitations significantly restrict the adaptability and practical application performance of the model.

@@

In this study, we focused on one particular direction: generalizing the Erlang distribution of stage durations currently present in the model. The goals were to have a more effective way for parameter estimation and to eliminate the constraints imposed by the shape parameter ($\kappa$). To address these limitations, we proposed the Geometric Chain Trick (GCT). Instead of having a variable number of substages $(n)$ with a constant transition rate between them $(\gamma_i = n\gamma)$, we fix $n$ to a specific integer value and use a series of rates that follow a geometric pattern. This series is represented by the formula $\gamma_i = ar^{i-1}$, governed by parameters $a$ and $r$. This approach enables a uniform model structure with the parameter of interest transitioning from discrete values $(n$ and $n\gamma)$ to continuous values $(a$ and $r)$, thereby enhancing the model's adaptability (details in Section 4.x).

@@

\section{Method}
Several assumptions have been made for the unforced SIR model (Susceptible, Infectious, and Recovered) that our study is based on: 
\begin{enumerate}
    \item \textbf{Closed Population}: A constant population size with equal birth and death rates and no disease-related mortalities.
    \item \textbf{Homogeneous mixing}: Individuals have an equal chance of coming into contact with any other individual in the population. 
    \item \textbf{Ignoring Latency}: Individuals move immediately from a susceptible state to an infected state, ignoring any incubation period associated with the disease.
    \item \textbf{Lifelong immunity}: Once an individual recovers, they do not become susceptible or infectious again.
    \item \textbf{Fixed rate of transmission and recovery}: The rate of disease spread and the rate at which infected individuals recover and become immune are constant over time.
\end{enumerate}

@@

\section{Results}
\subsection{Independent and Flexible Modification of Mean and Kappa in Stage Duration Distribution Using the GCT Model}
As discussed in previous sections, the duration distribution of the stage can be parameterized by its mean (D) and shape parameter $(\kappa^2 = CV^2)$. In the GCT model, the calculation of mean and variance is achieved by summing the respective properties of each substage, shown as follows:
\begin{align}
    Mean: \quad  &\sum_{i=1}^{n_f} \frac{1}{\gamma_i} = \frac{\frac{1}{a} (\frac{1}{r^n} - 1)}{\frac{1}{r} -1} \\
    Variance: \quad &\sum_{i=1}^{n_f} \left(\frac{1}{\gamma_i} \right)^2 = \frac{\frac{1}{a^2} (\frac{1}{r^{2n}} - 1)}{\frac{1}{r^2} -1}
\end{align}
The shape parameter $\kappa$, determined by $\frac{\sigma^2}{D}$, can thus be obtained:
\begin{align}
    \kappa = \frac{(1+r^{n}) (1-r)}{(1+r) (1-r^{n})}
\end{align}
This simplification revealed that the value of $\kappa$ is solely determined by the parameter $r$, thus enabling the alignment of kappa without affecting the mean, which can be adjusted through alterations to parameter $a$. Similarly, the mean can be scaled to a desired value without impacting $\kappa$. This grants the model independence in modifying properties.

@@

There are potential challenges that need to be addressed in the current GCT model. As discussed in Section 4.5, the parameters estimated after data fitting do not precisely reflect the actual values, yet these values are often crucial for understanding disease dynamics. For example, the basic reproduction number, $R_0$, which indicates how many secondary infections can be expected from a single infected individual in a fully susceptible population, is essential for assessing the potential for disease outbreak and epidemic spread. The calculation of $R_0$ requires the use of the contact rate $(\beta)$, recovery rate $(\gamma)$, and birth/death rate $(\mu)$. Therefore, achieving accurate parameter values from model fitting is necessary, and more study is needed.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{color}

@@

\lhead{Ningrui Xie}
\chead{Thesis First Draft}
\rhead{ID: 400323416}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0pt}

\newcommand{\perday}{\ensuremath{\mathrm{/day}}}
\newcommand{\Eqnref}[1]{Eq~\ref{#1}}
\newcommand{\eqnref}[1]{Eq~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}

\newcommand{\comment}{\showcomment}
%% \newcommand{\comment}{\nocomment}

\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}

@@

\newpage
\section{Abstract}
Simple dynamical models based on Ordinary Differential Equations (ODE) are widely utilized to forecast the spread of infectious diseases and to analyze other biological systems. A limitation of this framework is its implicit assumption that residence time in any given state, such as the infectious state, follows an exponentially distributed pattern. This assumption is often not realistic. A common solution is the application of the Linear Chain Trick (LCT), wherein a single stage is subdivided into multiple substages, each of which follows an identical exponential distribution, thereby yielding a more realistic Erlang distribution for the duration. However, the LCT poses challenges by restricting parametric flexibility, particularly resulting in less flexible dwell-time distributions, and it often necessitates comparisons between models with varying structures (i.e. different numbers of substages). Here, we propose a Geometric Chain Trick (GCT) employing a fixed number of substages with transition rates forming a geometric sequence. Through mathematical analysis, simulation, and model fitting, we demonstrate that our approach retains the principal benefits of the LCT while providing increased flexibility and facilitating easier data fitting. Compared to the recently introduced Generalized Linear Chain Trick (GLCT), which incorporates higher-dimensional auxiliary stages to model delays in transitions between main compartments, our method offers a more direct means of achieving flexibility. Although this study focuses on the application of the LCT and GCT in the epidemic SIR (susceptible-infectious-recovered) model, the potential applications of our approach span a broad array of dynamical models, including those for infectious diseases and other systems (e.g. population dynamics, biochemical reaction networks), offering advantages in computational efficiency, ease of parameter estimation, and the flexibility of time distributions.

\section{Introduction}
Ordinary differential equations (ODEs) are extensively utilized for mathematical modeling in the biological sciences, offering a robust framework to describe the temporal dynamics of systems. Their capacity to model continuous state changes makes them essential in capturing complex biological phenomena such as population dynamics \cite{salisbury2011mathematical}\cite{sego2021generation}, epidemiological patterns of infectious diseases \cite{Anderson1991}\cite{Diekmann2000}\cite{Feng2016}, and cellular processes like proliferation in oncogenesis \cite{baker1998modelling}\cite{jarrett2018mathematical}. The inherent flexibility and simplicity of ODEs make them not only easy to solve and understand but also easy to manipulate and accommodate to various situations.

In the field of epidemiology, ordinary differential equations (ODEs) serve as a crucial tool for characterizing disease progression and making predictions. During the COVID-19 pandemic, they were extensively utilized to guide policy-making, especially in disease control and public health interventions \cite{thompson2020epidemiological}. A commonly used model is the SIR compartmental model \cite{Anderson1991}\cite{kermack1927contribution}, which divides the population into three states: Susceptible $(S)$, Infectious $(I)$, and Recovered $(R)$. In the model, a susceptible individual, upon contact with the infected population at a rate $\beta$ $[\frac{1}{\mathit{ time}}]$, transitions immediately to the infectious state. An infected individual then moves to the recovered state at a recovery rate of $\gamma$ $[\frac{1}{\mathit{time}}]$. A major limitation of this model is its implicit assumption that the duration of the infectious stage follows an exponential distribution. This can be observed in a simplified equation that models the dynamics of an infected cohort. The proportion of a cohort infected at the same time begins at 1, and follows:
\begin{align}
    \frac{dI}{dt} = -\gamma I
\end{align}
Solving this equation explicitly yields
\begin{align}
    I(t) = e^{-\gamma t}.
	 \label{explicitCohort}
\end{align}
\Eqnref{explicitCohort} implies a stochastic state transition model where the time spend in the infectious $I$ is exponentially distributed with a mean stage duration of $\frac{1}{\gamma}$. \jd{Not really. Where did the stochastic part come from? Also, these things are implied by the assumptions, not the calculation.}
\Eqnref{explicitCohort} implies that an individual has the highest probability of recovering from the disease immediately after infection, which is generally not realistic. In reality, a delay between the successful infection of an individual and the recovery from the disease is inevitable. The work conducted by Byrne AW et al.~\cite{byrne2020inferred} presents the simulation of parameter distribution inferred for the duration of the infectious period, and the shape of this distribution closely resembles a gamma distribution. The study by Greenhalgh and Day (2017) \cite{greenhalgh2017time}, utilizing measles outbreak data, also demonstrates that the recovery rate from an infectious disease is not constant but can vary based on a wide range of factors. \jd{We're not really interested in factors, though, only in time-dependence. Is this citation good for that?}

Even when the mean is constant, the shape of the stage duration distributions (i.e., the timing of stage transitions) can significantly impact the dynamics and the outcomes of predictive models in applied settings \cite{krylova2013effects}\cite{keeling2002understanding}\cite{wearing2005appropriate}\cite{nguyen2008noise}. Consequently, it becomes crucial to consider the type of distribution characterizing an individual in a specific state when constructing these models.\jd{Try to be more consistent (and simpler) about what you call these distributions. Dwell-time, characterizing individual in state, stage-duration, are all saying the same thing.} A standard approach for developing models that incorporate different stage duration distributions is through the use of integro-differential equations (IDEs) or integral equations (IEs) \cite{hurtado2019generalizations}\cite{kermack1927contribution}\cite{hethcote1980integral}. IDEs facilitate the extension of the SIR model to include arbitrary distributions for the duration of infectiousness, as discussed in various sources \cite{feng2000endemic}\cite{hethcote1980integral}\cite{ma2006generality}. This approach allows for the capture of the `memory effect', where the current rate of change in each compartment of the model (Susceptible, Infectious, Recovered) is influenced by the entire history of the epidemic, rather than relying on a constant rate assumption.

@@

When applying LCT into the SIR model, the number of substages $n$ determines the shape of the stage duration distribution. As $n$ increases, the variance decreases, resulting in a distribution that is more peaked and narrow. As $n$ approaches infinity, the distribution converges to a delta function (a deterministic value, implying that all individuals who become infectious at time $t$ recover at exactly time $t + \frac{1}{\gamma}$), and the system approaches a delay differential equation \cite{krylova2013effects}\cite{hethcote1980integral}.

The Linear Chain Trick (LCT) is extensively employed in modeling biological processes, not only in quantifying disease progression \cite{lloyd2001destabilization}\cite{lloyd2001realistic}, but also in analyzing virus \cite{lloyd2001dependence}\cite{kakizoe2015method} and population \cite{cushing2013integrodifferential} dynamics. Studies have demonstrated that switching the stage duration distribution from an exponential to an Erlang distribution leads to predictions that align more closely with real data points \cite{kakizoe2015method}. But employing LCT in ODE models presents important limitations. First is inflexibility: the number of substages, $n$, must be an integer. Another limitation is the cumbersome and time-consuming process of determining the appropriate number of substages to use. Multiple models with varying substages must be constructed and fitted to identify the optimized parameter value. Prior research has investigated approaches for estimating the number of states in the linear chain, employing techniques such as profile likelihood \cite{raue2009structural} and model reduction \cite{maiwald2016driving}. A comparative study of these methods indicates that the typical quality of experimental data is insufficient for reliably estimating the chain length using these approaches \cite{hauber2020estimating}. These limitations significantly restrict the adaptability and practical application performance of the model.

@@

In this study, we focused on one particular direction: generalizing the Erlang distribution of stage durations currently present in the model. The goals were to have a more effective way for parameter estimation and to eliminate the constraints imposed on the shape parameter ($\kappa$). To address these limitations, we proposed the Geometric Chain Trick (GCT). Instead of having a variable number of substages $(n)$ with a constant transition rate between them $(\gamma_i = n\gamma)$, we fix $n$ to a specific integer value and use a series of rates that follow a geometric pattern. This series is represented by the formula $\gamma_i = ar^{i-1}$, governed by parameters $a$ and $r$. This approach enables a uniform model structure with the parameter of interest transitioning from discrete values $(n$ and $n\gamma)$ to continuous values $(a$ and $r)$, thereby enhancing the model's adaptability (details in \sref{Methods}).

@@

\section{Methods}
\label{Methods}
Several assumptions have been made for the unforced SIR model (Susceptible, Infectious, and Recovered) that our study is based on: 
\begin{enumerate}
    \item \textbf{Closed Population}: A constant population size with equal birth and death rates and no disease-related mortalities.
    \item \textbf{Homogeneous mixing}: Individuals have an equal chance of coming into contact with any other individual in the population. 
    \item \textbf{Ignoring Latency}: Individuals move immediately from a susceptible state to an infected state, ignoring any incubation period associated with the disease.
    \item \textbf{Lifelong immunity}: Once an individual recovers, they do not become susceptible or infectious again.
    \item \textbf{Fixed rate of transmission and recovery}: The rate of disease spread and the rate at which infected individuals recover and become immune are constant over time.
\end{enumerate}

@@

\section{Results}
\subsection{Independent and Flexible Modification of Mean and Kappa in Stage Duration Distribution Using the GCT Model}
\jd{What is new in this subsection? Can it be merged into Methods?}
As discussed in previous sections, the duration distribution of the stage can be parameterized by its mean (D) and shape parameter $(\kappa^2 = CV^2)$. In the GCT model, the calculation of mean and variance is achieved by summing the respective properties of each substage, shown as follows:
\begin{align}
    Mean: \quad  &\sum_{i=1}^{n_f} \frac{1}{\gamma_i} = \frac{\frac{1}{a} (\frac{1}{r^n} - 1)}{\frac{1}{r} -1} \\
    Variance: \quad &\sum_{i=1}^{n_f} \left(\frac{1}{\gamma_i} \right)^2 = \frac{\frac{1}{a^2} (\frac{1}{r^{2n}} - 1)}{\frac{1}{r^2} -1}
\end{align}
The shape parameter $\kappa$, determined by $\frac{\sigma^2}{D}$, can thus be obtained:
\begin{align}
    \kappa = \frac{(1+r^{n}) (1-r)}{(1+r) (1-r^{n})}
\end{align}
This simplification revealed that the value of $\kappa$ is solely determined by the parameter $r$, thus enabling the alignment of kappa without affecting the mean, which can be adjusted through alterations to parameter $a$. Similarly, the mean can be scaled to a desired value without impacting $\kappa$. This grants the model independence in modifying properties.

@@

\jd{Try to get the habit of doing referencing in latex instead of typing section numbers (see the example I tried with sref)}
There are potential challenges that need to be addressed in the current GCT model. As discussed in Section 4.5, the parameters estimated after data fitting do not precisely reflect the actual values, yet these values are often crucial for understanding disease dynamics. For example, the basic reproduction number, $R_0$, which indicates how many secondary infections can be expected from a single infected individual in a fully susceptible population, is essential for assessing the potential for disease outbreak and epidemic spread. The calculation of $R_0$ requires the use of the contact rate $(\beta)$, recovery rate $(\gamma)$, and birth/death rate $(\mu)$. Therefore, achieving accurate parameter values from model fitting is necessary, and more study is needed.
